Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
spell check

\begin{abstract}
The R programming language is widely used in a variety of domains for tasks
related to data science. R was designed to favor an interactive style of
programming with minimal syntactic and conceptual overhead. This design is
well suited to interactive data analysis, but a bad firt for tools such as
compilers or program analyzers which must generate performant code or catch
programming errors.  In particular, R has no type annotations, its
operations are dynamically checked at runtime. The starting point for our
work are the twin questions, \emph{what expressive power is needed to
  accurately type R code?} and \emph{which type system is the R community
  willing to adopt?} Both questions are difficult to answer without actually
designing a type system and attempting to convince users and package
developers to try the proposed system.  The goal of this paper is to provide
data that can feed into the design process. To this end, we perform a large
corpus analysis with aim to gain insights in the degree of polymorphism
exhibited by idiomatic R code and explore potential benefits that the R
community could accrue from even a simple type system.  As a starting point,
we infer type signatures for \NUMFUNCTIONS functions from \NUMPACKAGES
packages among the most widely used open source R libraries. The signatures
use a minimal type language with only basic types, vectors and classes.  We
implement a contract checker that can be used to validate the inferred type
signatures.
\end{abstract} \maketitle

@@

To sum up, our paper makes the following main contributions:
\begin{itemize}
\item We implemented tooling to automatically extract type signatures from R
  functions and to intrument R functions with checks based on their declared
  types. The tooling is robust and scalable to the entire R language.
\item We carried out a large-scale analyis of corpus of \PACKAGES widely
  used and actively maintained libraries to extract function type signatures
  validated the robustness of the inferred type signatures against XXX
  programs that use those functions.
\item We report on the appropriateness and usefulness of a simpe type
  language for the R programming language.
\end{itemize}

@@

But what if the design of the type system is unclear?  \citet{tip} propose
an intriguing approach called trace typing. With trace typing, a new type
system can be prototyped and evaluated by applying the type rule to
execution traces of programs. While the appraoch has the limitation of
dynamic analysis techniques, namely that the results are only as good at the
coverage of the source code, it allows to quickly test new design and
quantify how much of a code base can be type-checked.  Other approaches that
infer types for dynamic analysis include the work of \citet{FurrAF2009} for
Ruby.

There is no previous work for the R programming language. We take
inspiration in the above mentionned works but focus on adapting them to our
target language.

@@

Scalar data types include the distinguished \k{NULL} value, which is also of
ktype \k{NULL}, instance of classes written using the S4 object system,
closures and environments.  The implementation of R has a number of other
types listed in Figure~\ref{rtypes} for reference.

@@

Fig.~\ref{types} presents our type language. An early design choice was to
stay close to the R language and only depart in small, hopefully,
non-controverwsial ways.  Functions types have the form \FUN{A_1, \dots,
  A_n}\T where each $\A_i$ argument is either a type \T or \VARGS, a
variable length argument list. The rest of this section details and
motivates our design.

@@

\paragraph{Lists} Heterogeneous collections are implemented using lists.
Lists and vectors are closely related. Any vector can be converted to a list
with \code{as.list}, and lists to vectors with \code{unlist} (coercions may
ensue). We support parametrization of the list type, \LIST\T, but for
heterogenous lists this type will default to \any.

@@

\noindent
To obtain raw types, we make use of R's C FFI and use low-level machinery to
collect type tags and attributes from the R runtime.  The types that we
provide to users are constructed during post-processing, and rely on the
detailed information made available by these low-level reflection
mechanisms.

@@

The output of the \typetracer tool consists of a set of tuples of raw types,
each representing a function call in some program's execution. The inference
step consolidates the different tuples corresponding to a particular function
definition across multiple programs and distills them into a single type
signature.  The shape of the infered function signatures is:

@@

While \contractr's primary logic has been implemented in C++ to reduce its
runtime overhead, we have not observed a single segfault during our use of
the package.  It has been tested with GNU R-3.5.0 and hardened with a
battery of 400 test cases. It works by modifying function definition to
insert a call to a type-checking function.

In terms of useability, our checker is enabled automatically whenever a
package is loaded in R. So, a simple invocation of \code{library(contractr)}
causes contracts to be injected in all packages.  On loading, \contractr
scans all packages in the user's workspace and inserts contracts in
functions for which type signatures are available. Then \contractr sets up
package load hooks which are executed when new packages are loaded.
Furthermore, \contractr automatically removes contracts from all functions
and restores them to their original state when it is unloaded.

@@

\paragraph{Checking the values} A number of properties can be checked
by a simple tag check, namely wether a value is a: null, environment, vector
or scalar. Other properties require an inspection of the contents of a
value: the absence of \NA or the type of the elements of list. Union types
require checking for each of the alternatives. We do not check higher-order
functions in this version of the tool.

@@

These packages come from the Comprehensive R Achive Network
(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
code with over \AllCranRnd packages\footnote{CRAN receives about 6 new package
  submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd and
\AllNativeCodeRnd R and native code respectively. Unlike other open code
repositories such as GitHub, CRAN is a curated repository. Each submitted
package must abide to a number of well-formedness rules that are automatically
checked asserting certain quality. Most relevant for this work is that all of
the runnable code (including code snippets from examples and vignettes) is
tested and only a successfully running package is admitted in the archive.

@@

The first part of our evaluation attempts to shed light on how good a fit
our proposed type system is with respect to common programming patterns
occuring in widely used R libraries.

First we look at the share of monomorphic arguments and function signatures.
Monomorphic in this context means that the type is not relying on \any or
inlcuding a union.  The import of monorphism in this context is that it
means our type language can accurately capture an argument's type or a
function signature.

>>>>

\begin{abstract}
The R programming language is widely used in a variety of domains for tasks
related to data science. R was designed to favor an interactive style of
programming with minimal syntactic and conceptual overhead. This design is
well suited to interactive data analysis, but a bad fit for tools such as
compilers or program analyzers which must generate performant code or catch
programming errors.  In particular, R has no type annotations, its
operations are dynamically checked at run-time. The starting point for our
work are the twin questions, \emph{what expressive power is needed to
  accurately type R code?} and \emph{which type system is the R community
  willing to adopt?} Both questions are difficult to answer without actually
designing a type system and attempting to convince users and package
developers to try the proposed system.  The goal of this paper is to provide
data that can feed into the design process. To this end, we perform a large
corpus analysis with aim to gain insights in the degree of polymorphism
exhibited by idiomatic R code and explore potential benefits that the R
community could accrue from even a simple type system.  As a starting point,
we infer type signatures for \NUMFUNCTIONS functions from \NUMPACKAGES
packages among the most widely used open source R libraries. The signatures
use a minimal type language with only basic types, vectors and classes.  We
implement a contract checker that can be used to validate the inferred type
signatures.
\end{abstract} \maketitle

@@

To sum up, our paper makes the following main contributions:
\begin{itemize}
\item We implemented tooling to automatically extract type signatures from R
  functions and to instrument R functions with checks based on their declared
  types. The tooling is robust and scalable to the entire R language.
\item We carried out a large-scale analysis of corpus of \PACKAGES widely
  used and actively maintained libraries to extract function type signatures
  validated the robustness of the inferred type signatures against XXX
  programs that use those functions.
\item We report on the appropriateness and usefulness of a simple type
  language for the R programming language.
\end{itemize}

@@

But what if the design of the type system is unclear?  \citet{tip} propose
an intriguing approach called trace typing. With trace typing, a new type
system can be prototyped and evaluated by applying the type rule to
execution traces of programs. While the approach has the limitation of
dynamic analysis techniques, namely that the results are only as good at the
coverage of the source code, it allows to quickly test new design and
quantify how much of a code base can be type-checked.  Other approaches that
infer types for dynamic analysis include the work of \citet{FurrAF2009} for
Ruby.

There is no previous work for the R programming language. We take
inspiration in the above mentioned works but focus on adapting them to our
target language.

@@

Scalar data types include the distinguished \k{NULL} value, which is also of
type \k{NULL}, instance of classes written using the S4 object system,
closures and environments.  The implementation of R has a number of other
types listed in Figure~\ref{rtypes} for reference.

@@

Fig.~\ref{types} presents our type language. An early design choice was to
stay close to the R language and only depart in small, hopefully,
non-controversial ways.  Functions types have the form \FUN{A_1, \dots,
  A_n}\T where each $\A_i$ argument is either a type \T or \VARGS, a
variable length argument list. The rest of this section details and
motivates our design.

@@

\paragraph{Lists} Heterogeneous collections are implemented using lists.
Lists and vectors are closely related. Any vector can be converted to a list
with \code{as.list}, and lists to vectors with \code{unlist} (coercions may
ensue). We support parametrization of the list type, \LIST\T, but for
heterogeneous lists this type will default to \any.

@@

\noindent
To obtain raw types, we make use of R's C FFI and use low-level machinery to
collect type tags and attributes from the R run-time.  The types that we
provide to users are constructed during post-processing, and rely on the
detailed information made available by these low-level reflection
mechanisms.

@@

The output of the \typetracer tool consists of a set of tuples of raw types,
each representing a function call in some program's execution. The inference
step consolidates the different tuples corresponding to a particular function
definition across multiple programs and distills them into a single type
signature.  The shape of the inferred function signatures is:

@@

While \contractr's primary logic has been implemented in C++ to reduce its
runtier overhead, we have not observed a single segfault during our use of
the package.  It has been tested with GNU R-3.5.0 and hardened with a
battery of 400 test cases. It works by modifying function definition to
insert a call to a type-checking function.

In terms of usability, our checker is enabled automatically whenever a
package is loaded in R. So, a simple invocation of \code{library(contractr)}
causes contracts to be injected in all packages.  On loading, \contractr
scans all packages in the user's workspace and inserts contracts in
functions for which type signatures are available. Then \contractr sets up
package load hooks which are executed when new packages are loaded.
Furthermore, \contractr automatically removes contracts from all functions
and restores them to their original state when it is unloaded.

@@

\paragraph{Checking the values} A number of properties can be checked
by a simple tag check, namely whether a value is a: null, environment, vector
or scalar. Other properties require an inspection of the contents of a
value: the absence of \NA or the type of the elements of list. Union types
require checking for each of the alternatives. We do not check higher-order
functions in this version of the tool.

@@

These packages come from the Comprehensive R Archive Network
(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
code with over \AllCranRnd packages\footnote{CRAN receives about 6 new package
  submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd and
\AllNativeCodeRnd R and native code respectively. Unlike other open code
repositories such as GitHub, CRAN is a curated repository. Each submitted
package must abide to a number of well-formedness rules that are automatically
checked asserting certain quality. Most relevant for this work is that all of
the runnable code (including code snippets from examples and vignettes) is
tested and only a successfully running package is admitted in the archive.

@@

The first part of our evaluation attempts to shed light on how good a fit
our proposed type system is with respect to common programming patterns
occurring in widely used R libraries.

First we look at the share of monomorphic arguments and function signatures.
Monomorphic in this context means that the type is not relying on \any or
including a union.  The import of monorphism in this context is that it
means our type language can accurately capture an argument's type or a
function signature.
